---
title: "Derivation of some of the orginal paper results"
subtitle: "Supporting material for Standardized means difference effect size measures for planned comparisons, trend analysis and other applications of contrast analysis"
author: "mc & mp"
date: '`r format(Sys.Date(), "%m, %d, %Y")`'
bibliography: contrasts.bib
header-includes:
   - \usepackage{amsmath}
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE}
library("papaja")
knitr::opts_chunk$set(echo = TRUE)
source("resources/writing_functions.R")
knit_refs()
ref_paper<-"smdeffectsizepaper"

```

# Aim

Here we derive some of the most important formula in the paper. In this document, equations referred to with REF. # indicates the reference number of the formula in the original paper. 

# Contrasts analysis background

A contrast is a linear combination of means whose coefficients sum up to zero, meant to estimate a particular comparison of means and test it against zero. We refer to the contrast set of coefficients as $\boldsymbol{c}=\{c_i\}$, and to the expected set of means as $\boldsymbol{\mu}=\{\mu_i\}$. The contrast coefficients (weights) are chosen such that $\sum_i{c_i}=0$, with $i=\{1,..,k\}$ where $k$ is the number of means being compared. The contrast expected value is $c\mu=\sum_i{(c_i \cdot \mu_i)}$. As an example, consider a simple design with two groups: the comparison of the two groups means can be carried out with a simple contrast with $\boldsymbol{c}=\{1,-1\}$, in which the contrast value is simply the expected difference between means, $c\mu=c_1\mu_1+c_2\mu_2=\mu_1-\mu_2$.  

A contrast defined across $k$ means of independent groups of size $n$ can be tested employing either an independent samples t-test or an F-test. The t-test expected value, with $k(n-1)$ degrees of freedom, is [@steiger2004beyond, p.172 EQ 43 and p.173, EQ 46]:

$$ E(t_{k(n-1)})={\sum{(c_i \cdot \mu_i)} \over {\sigma \cdot  \sqrt{\sum{c_i^2} \over n}}} \tag{`r ref_tag("et")`}$$

The F-test associated with a contrast is simply $F_{1,k(n-1)}=t_{k(n-1)}^2$.

# Cohen's $\delta$ measures for contrasts

@cohen1988 defines several indices of effect size for the comparison of two means. In the context of two-groups designs, he defines:
$$\delta={{\mu_1-\mu_2} \over \sigma} \tag{`r ref_tag("delta")`}$$ 
When the same logic is applied to a contrast comparison, it naturally generalizes to [cf. @steiger2004beyond] 

$$\delta_0={\sum{(c_i \cdot \mu_i)} \over \sigma} \tag{`r ref_tag("delta0")`}$$ 

 @steiger2004beyond, pag. 173, EQ 46, defines
 
 $$E_s={\sum{c_i\mu_i} \over {\sigma}} \tag{`r eq("staiger1")`}$$
which is $\delta_0$.

## Standardized contrast effect size measure

$$\delta_z={\sum{(c_i \cdot \mu_i)} \over {\sigma \sqrt{\sum{c_i^2}}}} \tag{`r ref_tag("dz")`}$$
If we set $z={1/\sqrt{\sum{c_i^2}}}$, we obtain:
$$\delta_z={z \cdot \delta_0} \tag{`r ref_tag("dzshort")`}$$

### Result in the paper
$$f={{1 \over {\sqrt{k}} } \cdot \delta_z} \tag{`r ref_tag("ztof")`}$$

### Derivation

@cohen1988, p. 274 [@maxwell1990designing, EQ. 89] defines $f$ as the ratio between the expected standard deviation of the means and the error standard deviation:

$$f={\sigma_\mu \over \sigma}$$
in our case, we are looking for:
$$f={\sigma_\psi \over \sigma}$$
where $\psi$ is any contrast of interest. 
@gpower, p. 29 defines 
$$\hat{\sigma_{\psi}}={\sqrt{{(\sum{c_i\hat{\mu_i}})^2\over N{\sum{c_i^2} \over n}}}}$$
since $N=kn$, we have
$$\sigma_{\psi}={\sqrt{{(\sum{c_i\mu_i})^2\over k\sum{c_i^2}}}}$$
thus,
$$\sigma_{\psi}={\sum{c_i\mu_i}\over\sqrt{ k\sum{c_i^2}}}$$
and 
$$f={\sigma_{\psi} \over \sigma}={\sum{c_i\mu_i}\over\sqrt{ k\sum{c_i^2}}}={{1 \over \sqrt{k}} \cdot \delta_z}$$
$\hfill\blacksquare$

However, it is interesting to show the same result without dealing with sample estimates. Consider a regression with $k-1$ contrasts, one is the contrast of interest $\psi$, and the others, $\phi_i$, are contrasts orthogonal to each other and to the contrast of interest.  
The total variance of the $y$ variable will be:
$$\sigma_t^2=\sigma_\psi^2+\sum{\sigma_{\psi_i}^2}+\sigma^2$$
because of the orthogonality of the contrasts. The population $b$ coefficient associated with $\psi$ is:
$$b_\psi={\sum{c_i\mu_i} /k \over \sum{c_i^2}/k}={\sum{c_i\mu_i} \over \sum{c_i^2}}$$
and the variance examplained by the contrast $\psi$ is given by:
$$\sigma_\psi^2=b_\psi^2 \cdot \sigma_c^2={(\sum{c_i\mu_i})^2 \over (\sum{c_i^2})^2} \cdot {\sum{c_i^2} \over k}= {(\sum{c_i\mu_i})^2 \over \sum{c_i^2}} \cdot {1 \over k}$$
Taking the square root yields:
$$\sigma_\psi= {1 \over \sqrt{k}} \cdot {\sum{c_i\mu_i} \over \sqrt{\sum{c_i^2}}}$$
because we were looking for the $f=\sigma_\psi/\sigma$ ratio, we obtain
$$f={\sigma_\psi \over \sigma}={1 \over \sqrt{k}} \cdot {\sum{c_i\mu_i} \over \sigma \sqrt{\sum{c_i^2}}}={1 \over \sqrt{k}} \cdot \delta_z$$
$\hfill\blacksquare$


```{r, include=FALSE}
# cohen's examples pag 279
#pattern 1
d<-1.3
k<-7
m<-c(-.5*d,0,0,0,0,0,.5*d)
cc<-c(-1,0,0,0,0,0,1)
d*sqrt(1/(2*k))
(dz<-sum(m*cc)/sqrt(sum(cc^2)))
dz/sqrt(k)

#pattern 2
d<-2
k<-7
m<-c(-3*d/(k-1),-2*d/(k-1),-d/(k-1),0,d/(k-1),2*d/(k-1),3*d/(k-1))*48
cc<-c(-3,-2,-1,0,1,2,3)
.5*d*sqrt((k+1)/(3*(k-1)))
(dz<-sum(m*cc)/sqrt(sum(cc^2)))
sum(m*cc)/sqrt(sum(cc^2))

dz/sqrt(k)
### alternative 
cf<-rep(cc,each=10000)
vc<-var(cf)*(length(cf)-1)/length(cf)
(sum(cc^2)/k)

cf<-factor(rep(cc,each=10000))
contrasts(cf)<-contr.poly(7)
y<-rep(m,each=10000)+rnorm(k*10000)
z<-as.numeric(scale(y))
lm(y~cf)
sum(m*cc)/sum(cc^2)


sqrt(var(y))

apply(contrasts(cf),2,function(a) sum((a)^2))

```
### Paper result

$$\eta^2={\delta_z^2 \over {\delta_z^2+k}}\tag{`r ref_tag("eta")`}$$

### Derivation


$$\eta^2={\sigma_\psi^2 \over {\sigma_\psi^2+\sigma^2}}$$
we have shonw that
$$\sigma_\psi^2= {1 \over k} \cdot {(\sum{c_i\mu_i})^2 \over {\sum{c_i^2}}}$$
thus
$$\eta^2={{1 \over k} \cdot {(\sum{c_i\mu_i})^2 \over {\sum{c_i^2}}} \over {{1 \over k} \cdot {(\sum{c_i\mu_i})^2 \over {\sum{c_i^2}}}+\sigma^2}}=$$
$$={{1 \over k} \cdot {(\sum{c_i\mu_i})^2 \over {\sum{c_i^2}}} \over {(\sum{c_i\mu_i})^2+k\sigma^2\sum{c_i^2}  \over{k\sum{c_i^2}}}}=$$
$$={{(\sum{c_i\mu_i})^2 } \over {(\sum{c_i\mu_i})^2+k\sigma^2 \sum{c_i^2} }}=$$
let's devide numerator and denominator by $\sigma^2 \sum{c_i^2}$
$$={{(\sum{c_i\mu_i})^2 \over \sigma^2 \sum{c_i^2} } \over {{(\sum{c_i\mu_i})^2 \over \sigma \sum{c_i^2}}+{k\sigma^2 \sum{c_i^2} \over \sigma^2 \sum{c_i^2}}}}=$$
$$={{\delta_z^2} \over {\delta_z^2+k}}$$
$\hfill\blacksquare$

## Scaled effect size measure

A different method of scaling the constrat effect size measure which guarantees better interpretability and comparability can be suggested. Let's $g={2 \over \sum_i{\left|{c_i}\right|}}$, where $|c_i|$ indicates the absolute value of $c_i$, then  

$$\delta_g=g \cdot \delta_0={2 \over \sum{|{c_i}|}} \cdot {{\sum_i{c_i \cdot \mu_i}} \over  {\sigma }}  \tag{`r ref_tag("dg")`}$$

### Paper results

All the derivations for $\delta_z$ can be derived for $\delta_g$ by noticing that 

$$\delta_z={{z \over g} \delta_g}$$
thus, one simply substitutes ${{z \over g} \delta_g}$ is place of $\delta_z$. As an example, we derive here:

$$\eta_p^2={\delta_g^2 \over { \delta_g^2+({{g^2  \over z^2} \cdot k})}} \tag{`r ref_tag("gtoeta")`}$$

### Derivation

Recall that 
$$={{\delta_z^2} \over {\delta_z^2+k}}$$

and substitutes ${{z \over g} \delta_g}$ is place of $\delta_z$:

$$\eta_p^2={{z^2 \over g^2}{\delta_g^2} \over {{z^2 \over g^2}{\delta_g^2}+k}}=$$

$$={{z^2\delta_g^2} \over {z^2\delta_g^2+g^2k }}$$
we now divide numerator and denominator by $z^2$ and we obtain:
$$={{\delta_g^2} \over {\delta_g^2+ {g^2 \over z^2}k}}$$
$\hfill\blacksquare$

# Sample Estimation

$$d_z={\sum{(c_i \cdot m_i)} \over {s_p \sqrt{\sum{c_i^2}}}}  \tag{`r eq("sampledz")`}$$

and 

$$d_g=2 \cdot {\sum{(c_i \cdot m_i)} \over {s_p \sum{|{c_i}|}}}  \tag{`r eq("sampledg")`}$$

### Paper result

$$d_z =  {t_{k(n-1)} \over {\sqrt{ n}} }  \tag{`r ref_tag("samplettoz")`}$$

### Derivation

Recall that 
$$t_{k(n-1)}=\sqrt{n} \cdot d_z  \tag{`r ref_tag("ttodz")`}$$ 
thus
$$d_z =  {t_{k(n-1)} \over {\sqrt{ n}} }$$

$\hfill\blacksquare$

### Paper result

$${\delta_z}={\sqrt{k \cdot {n-1 \over n}} \cdot { r_{contrast} \over \sqrt{1-r_{contrast}^2}}} \tag{`r ref_tag("zfromr")`}$$

### Derivation

Following [@rosenthal2000contrasts], pag. 38, EQ 3.2

$$F_{contrast}={MS_{contrast}\over{MS_{within}}}$$
and 

$$r_{contrast}=\sqrt{F_{contrast} \over{F_{contrast}+k(n-1)}}$$
thus

$$r_{contrast}=\sqrt{t_{contrast}^2 \over{t_{contrast}^2+k(n-1)}}$$
therefore
$$r_{contrast}=\sqrt{n\cdot d_z^2 \over{n \cdot d_z^2+k(n-1)}}$$
by dividing num and den by $n$, we obtain:
$$r_{contrast}=\sqrt{d_z^2 \over{ d_z^2+{k(n-1)\over n}}}$$
which means:
$$r_{contrast}^2={d_z^2 \over{ d_z^2+{k(n-1)\over n}}}$$
$$r_{contrast}^2({ d_z^2+{k(n-1)\over n}})={d_z^2}$$
$${r_{contrast}^2 d_z^2 +{r_{contrast}^2{k(n-1)\over n}}}={d_z^2}$$
$${r_{contrast}^2{k(n-1)\over n}}={d_z^2}-{r_{contrast}^2 d_z^2}$$
$${r_{contrast}^2{k(n-1)\over n}}={d_z^2}(1-{r_{contrast}^2)}$$
$${{k(n-1)\over n} {\cdot {r_{contrast}^2} \over 1-{r_{contrast}^2}}} ={d_z^2}$$
$${d_z}=\sqrt{{k(n-1)\over n}} \cdot {{r_{contrast}} \over \sqrt{1-{r_{contrast}^2}}}$$
$\hfill\blacksquare$

# Power Analysis

Let $\lambda_k$ be the noncentrality parameter of the $t_{k(n-1)}$ distribution associated with the t-test for $d_z$.

### Result
$$\lambda_k={\sqrt{n} \cdot d_z} \tag{`r ref_tag("noncentdz")`}$$


Following @liu2013power, page 87, EQ 4.25,

$$\lambda_k={{\sum{c_i\mu_i}}\over{\sqrt{\sigma^2({\sum{c_i^2\over n_i}})}}}$$
which simplifies for equal cell size to:
$$\lambda_k={\sqrt{n} \cdot {{\sum{c_i\mu_i}}\over{\sqrt{\sigma^2 \cdot {\sum{c_i^2}}}}}}$$
and thus
$$\lambda_k={\sqrt{n} \cdot d_z}$$
$\hfill\blacksquare$

### Paper result

The non-centrality parameter of the F-test is given by: 
$$\hat{\lambda_F}=k\cdot n \cdot f^2 \tag{`r ref_tag("noncentf")`}$$


### Derivation

Note that this is not a result of the paper, since it is taken from @steiger2004beyond, p. 169 EQ 10 or @cohen1988, p. 481,  @steiger1997noncentrality, p. 246 EQ 9.95.  However, it is interesting to derive it from `r ref_tag("noncentdz")`.

First, notice that for contrasts with 1 numerator $df$, $\lambda_F=\lambda_t^2$. Therefore

$$\lambda_F={n \cdot d_z^2}$$
we can multiply and divide by $k$, and we get
$$\lambda_F={{n \cdot k \over k } \cdot d_z^2}$$
since $f=d_z/\sqrt{k}$, we reach:
$$\lambda_F={n \cdot k \cdot f^2}$$

$\hfill\blacksquare$

# Confidence intervals

All the formulas in this sections are derived directly from the literature. We only want to check that 
$$Pr\left[ {{2 \over \sum{|{c_i}|}} \cdot \sqrt{\sum{c_i^2} \over n}} \cdot \hat{\lambda_l} \le d_g \le {{2 \over \sum{|{c_i}|}} \cdot \sqrt{\sum{c_i^2} \over n}} \cdot \hat{\lambda_l} \right]=1-\alpha \tag{`r ref_tag("cipowerdg")`}$$
can be derived from 
$$Pr\left[{{\hat{\lambda_l} \over \sqrt{n}} \le d_z \le {\hat{\lambda_u} \over {\sqrt{n}}}  }\right]=1-\alpha \tag{`r ref_tag("power")`}$$
by substituing $d_z$ with $d_g$.

### Derivation

Recall that $d_g={g \over z}d_z$. Thus, for $d_g$ we have:
$$Pr\left[{{g \over z}{\hat{\lambda_l} \over \sqrt{n}} \le {{g \over z}d_z} \le {g \over z}{\hat{\lambda_u} \over {\sqrt{n}}}  }\right]=1-\alpha $$
$$Pr\left[{{g \cdot z^{-1}}{\hat{\lambda_l} \over \sqrt{n}} \le {d_g} \le {g \cdot  z^{-1}}{\hat{\lambda_u} \over {\sqrt{n}}}  }\right]=1-\alpha \tag{`r ref_tag("power")`}$$
$$Pr\left[{{{2 \over \sum{|c_i|}} \cdot z^{-1}}{\hat{\lambda_l} \over \sqrt{n}} \le {d_g} \le {{2 \over \sum{|c_i|}} \cdot  z^{-1}}{\hat{\lambda_u} \over {\sqrt{n}}}  }\right]=1-\alpha$$
$$Pr\left[{{{2 \over \sum{|c_i|}} \cdot \sqrt{\sum{c_i^2}}}{\hat{\lambda_l} \over \sqrt{n}} \le {d_g} \le {{2 \over \sum{|c_i|}} \cdot  \sqrt{\sum{c_i^2}}}{\hat{\lambda_u} \over {\sqrt{n}}}  }\right]=1-\alpha$$
$$Pr\left[ {{2 \over \sum{|{c_i}|}} \cdot \sqrt{\sum{c_i^2} \over n}} \cdot \hat{\lambda_l} \le d_g \le {{2 \over \sum{|{c_i}|}} \cdot \sqrt{\sum{c_i^2} \over n}} \cdot \hat{\lambda_l} \right]=1-\alpha $$
$\hfill\blacksquare$


# References
```{r create_r-references, echo=F}
r_refs(file = "contrasts.bib")

```
